---
title: "Untitled"
author: "Chris Peralta"
date: "September 15, 2018"
output: 
  pdf_document
    toc: true
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

library(nycflights13) # Available on CRAN, contains datasets
library(tidyverse)
```

```{r pressure, include=FALSE}
airlines <- airlines
airports <- airports
flights <- flights
planes <- planes
weather <- weather
```

In this project, I will attempt to predict departure delays using linear regression. I will begin by ...

Before I begin, I'll tell you a little about the data. 
- `airlines` is a list of airlines and their abbreviations
- `airports` is a list of airports with their locations, timezones, and faa codes
- `flights` is a list of all flights that departed NYC in 2013 with other related data
- `planes` is a dataset of all of the planes that went on the flights above
- `weather` is a dataset of the weather conditions by hour and airport for the year of 2013

Below are some summary statistics for the `flights` dataset. 
```{r dep_delay}
summary(flights)
```

There are a few interesting observations above: 
- `time_hour` has no clear meaning (it contains times rounded down to the nearest hour for joining the `weather` data with `flights`)
- `air_time` aappears to be in minutes
- Arrival and departure times are given in a 4-digit format

